Synthesis of Linear Functional Programs

### 2021-05-05 article/article.tex
minor change
more introduction

\usepackage[inline]{todo}
\usepackage{cmll}
\usepackage{mathpartir}

@@

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita}
\date{April 2021}
\institute{NOVA School of Science and Technology}

@@

This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
Said correspondence states that propositions in a logic have a direct mapping to types, and well typed programs also have a one to one correspendence to proofs of those propositions.
As such, having a type be a proposition in linear logic, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, and so we formulate the synthesis as proof search in linear logic.

\begin{enumerate}
\item General motivation
  \begin{itemize}
\item Program synthesis is an automated or semi-automated process of
  deriving a program from a specification of its behavior.
\item Specifications appear in many forms\dots
\item Program synthesis as a means of improving program
  correctness/programmer productivity.
\item Challenges: space of valid programs, space of valid
  specifications,  etc.
\end{itemize}
\item \emph{Type-based} program synthesis
  \begin{itemize}
 \item Rich types as specifications.
\item Languages with richer type systems allow for more precise
  types. This precision can statically eliminate various kinds of
  logical errors by making certain invalid program states ill-typed
  (e.g. a ``null aware'' type system, will ensure you cannot
  dereference a null-pointer).
\item But can also be a burden -- the whole point of these type
  systems is to ensure that ``less'' things are well-typed, and
  sometimes its hard to convince the checker\dots
\item Type-based synthesis leverages rich types as a way of pruning
  the search space and by using types as a more ``familiar''
  specification.
\end{itemize}
\item ``My'' problem: Synthesis based on linear types
  \begin{itemize}
  \item Linear types constrain resource usage in programs by \emph{statically} limiting
    the number of times certain resources can be used during its
    lifetime.
  \item Applications in resource-aware programming such as concurrent
    programming (e.g. session types) and memory-management
    (e.g. Rust).
  \item This work will study program synthesis in a functional,
    linearly-typed setting, by leveraging the propositions-as-types
    correspondence between linear logic and the (linear)
    $\lambda$-calculus.
  \item What is props-as-types? What is linear logic? Linear $\lambda$
    is ``just'' a small functional core language.
  \item Through props-as-types, synthesis can be formulated as proof
    search and so we can leverage a lot of related technology from the
    literature.
  \end{itemize}
\item Goals:
  \begin{itemize}
\item Full and partial synthesis of well-typed (in the above sense)
  programs. Full synthesis means producing a function (or set of)
  satisfying the spec. Partial synthesis means being given a partial
  program (i.e. a function with a hole in it) and ``completing it''.
\item Start from a small system, build up to recursive
  types/functions.
\item Potential to go further: type refinements; interactive
  synthesis?
 \item Evaluation through expressiveness benchmarks (can we synthesize
   ``X''?); time measurements (how fast can we synthesize ``X''?');
   etc.
  \end{itemize}

@@

\[
  \infer[$(\lolli\! I)$]
  {\Delta , x{:}A \vdash M : B }
  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
  \quad
  \infer[$(\lolli\! E)$]
  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
  {\Delta_1, \Delta_2 \vdash M\,N : B}
\]
\[
  \infer*[left=($\tensor I$)]
  {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
  {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
  \quad
  \infer*[right=($\tensor E$)]
  {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
    N : C }
  {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
\]

>>>>

\usepackage{cmll}
\usepackage{mathpartir}

@@

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita \and Bernardo Toninho}
\date{April 2021}
\institute{NOVA School of Science and Technology}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of
 deriving a program (generating code) from a high-level specification.
 Specifications appear in many forms (i.e.
 polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
 examples~\cite{DBLP:conf/popl/FrankleOWZ16},
 graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...),
 and in this work the synthesis specification will be a linear type, ?possibly with refinements.?
 The chosen specification form dictates the synthesis process, and different ones might
 be approached in completely different ways.
 In spite of the specification, most program synthetizer's have to deal with two main
 inherent complexities: searching over a vast space of valid programs, and interpreting user intent.
 Synthesis has many motivations, it can be seen as a means to improve a programmer's productivity and/or
 program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
 of the programming process (i.e. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
 than to replace), just to name a few examples.

% Type based synthesis
Type-based synthesis takes a given type and produces a program with that same type.

This work explores the process of synthetising linear functional
programs from types based in linear logic, leveraging the Curry-Howard correspondence.
Said correspondence states that propositions in a logic have a direct mapping to types,
 and well typed programs also have a one to one correspendence to proofs of those propositions.
As such, having a type be a proposition in linear logic, we can relate a proof of that
proposition directly to a linear functional program
 — finding a proof is finding a program with that type, and so we
 formulate the synthesis as proof search in linear logic.

\begin{enumerate}
\item General motivation
  \begin{itemize}
\item 
\item Specifications appear in many forms\dots
\item Program synthesis as a means of improving program
  correctness/programmer productivity.
\item Challenges: space of valid programs, space of valid
  specifications,  etc.
\end{itemize}
\item \emph{Type-based} program synthesis
  \begin{itemize}
 \item Rich types as specifications.
\item Languages with richer type systems allow for more precise
  types. This precision can statically eliminate various kinds of
  logical errors by making certain invalid program states ill-typed
  (e.g. a ``null aware'' type system, will ensure you cannot
  dereference a null-pointer).
\item But can also be a burden -- the whole point of these type
  systems is to ensure that ``less'' things are well-typed, and
  sometimes it's hard to convince the checker\dots
\item Type-based synthesis leverages rich types as a way of pruning
  the search space and by using types as a more ``familiar''
  specification.
\end{itemize}
\item ``My'' problem: Synthesis based on linear types
  \begin{itemize}
  \item Linear types constrain resource usage in programs by \emph{statically} limiting
    the number of times certain resources can be used during its
    lifetime.
  \item Applications in resource-aware programming such as concurrent
    programming (e.g. session types) and memory-management
    (e.g. Rust).
  \item This work will study program synthesis in a functional,
    linearly-typed setting, by leveraging the propositions-as-types
    correspondence between linear logic and the (linear)
    $\lambda$-calculus.
  \item What is props-as-types? What is linear logic? Linear $\lambda$
    is ``just'' a small functional core language.
  \item Through props-as-types, synthesis can be formulated as proof
    search and so we can leverage a lot of related technology from the
    literature.
  \end{itemize}
\item Goals:
  \begin{itemize}
\item Full and partial synthesis of well-typed (in the above sense)
  programs. Full synthesis means producing a function (or set of)
  satisfying the spec. Partial synthesis means being given a partial
  program (i.e. a function with a hole in it) and ``completing it''.
\item Start from a small system, build up to recursive
  types/functions.
\item Potential to go further: type refinements; interactive
  synthesis?
 \item Evaluation through expressiveness benchmarks (can we synthesize
   ``X''?); time measurements (how fast can we synthesize ``X''?');
   etc.
  \end{itemize}

@@

\[
  \infer[$(\lolli\! I)$]
  {\Delta , x{:}A \vdash M : B }
  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
  \quad
  \infer[$(\lolli\! E)$]
  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
  {\Delta_1, \Delta_2 \vdash M\,N : B}
\]

\[
  \infer*[left=($\tensor I$)]
  {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
  {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
  \quad
  \infer*[right=($\tensor E$)]
  {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
    N : C }
  {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
\]
